---
title: "Hands-on_Exercise 1"
author: "Liu Chih Yuan"
---

# Getting Started

## **Install and launching R packages**

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in this computer. If they are, then they will be launched into R.

*Pacman is like a package manager, Tidyverse are packages for data science eg. plotting, wrangling, cleaning, visualization etc*

```{r}
pacman::p_load(tidyverse)
```

## **Importing the data**

*Note if you did not run the first code chunk tidyverse, this will indicate not found. read_csv is a code from tidyverse*

Looks very much like what we are doing in Pandas pd.read()

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Ploting a simple bar chart

If you want to show the code chunk only: add "#\| eval: false"

If you want to show the plot only: add "#\| echo: false"

**Down below is a example of showing plot only**

*The best thing about R Markdown / Quarto is the ability to knit code + text + graph into full documents*

"Kint" means taking .Rmd or .qmd file (with code chunks, text, and plots) and convert into a polished document like:

1.  PDF

2.  HTML website

3.  Word documents

4.  Slides / presentations

```{r}
#| echo: false
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

Here is a code chunk that shows everything as default
```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```
