---
title: "Take-Home Exercise 1"
author: "Liu Chih Yuan"
date: "April 29, 2025"
date-modified: "last-modified"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Exploratory Analysis of Age and Gender Distribution Across Singapore Planning Areas (2024)

## 1. Overview

The 2024 dataset from Singapore' Department of Statistics provides resident population data by planning area, subzone, age, and sex.

### 1.1 Objective

To perform structured exploratory data analysis to uncover insights on demographic distribution across regions.

## 2. Getting Started

### 2.1 Load Packages

```{r}
pacman::p_load(tidyverse, ggrepel, ggthemes, patchwork, ggridges, scales)
```

The R packages used in this EDA are as follows:

1.  `tidyverse` core R package for data science (contains essential packages such as ggplot2)
2.  `ggrepel` for ggplot2 to repel overlapping text labels
3.  `ggthemes` extra ggplot themes
4.  `patchwork` combine ggplot
5.  `ggridges` for ridgeline plots
6.  `scales` customer number formatting

### 2.2 Import Data

```{r}
df <- read.csv("data/respopagesex2024.csv")
```

```{r}
common_theme <- theme_minimal(base_size = 16) +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 20),
    plot.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 17),
    legend.title = element_text(size = 16)
  )
```

#### 2.2.1 Check Missing Values

```{r}
glimpse(df)
colSums(is.na(df))
```

### 2.3 Data Processing

#### 2.3.1 Checking Column Types

```{r}
str(df)
```

We see column `Age` is "chr" (character) instead of numeric, let's check why by finding the unique value of column `Age`

```{r}
sort(unique(df$Age))
```

Here most likely it's the `90_and_over` causing it to be a `str` not `int`

#### 2.3.2 Creating a Numeric Age Column

```{r}
df <- df %>%
  mutate(
    AgeNum = suppressWarnings(
      ifelse(Age == "90_and_Over", 90, as.numeric(Age))
    )
  )
```

Since we observed people over 90 years old are categorized `90_and_above` instead of actual numbers, for the ease of plotting we hereby create a new column `AgeNum`

#### 2.3.3 Creating Age Grouping

```{r}
df <- df %>%
  mutate(
    AgeGroup = case_when(
      AgeNum <= 12 ~ "Child",
      AgeNum <= 24 ~ "Youth",
      AgeNum <= 64 ~ "Adult",
      TRUE ~ "Senior"
    )
  )
```

We create a new column `AgeGroup` for future EDA purposes

```{r}
str(df)
```

```{r}
common_theme <- theme_minimal(base_size = 14) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )
```

## 3. Exploratory Data Analysis

### EDA 1: Population by Age / Age Group

```{r fig.height=8, fig.width=10}
p1 <- ggplot(df, aes(x = AgeNum, y = Pop)) +
  stat_summary(fun = sum, geom = "bar", fill = "steelblue") +
  labs(title = "Total Population by Age", x = "Age", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

p2 <- df %>%
  group_by(AgeGroup) %>%
  summarise(Pop = sum(Pop)) %>%
  ggplot(aes(x = AgeGroup, y = Pop, fill = AgeGroup)) +
  geom_bar(stat = "identity") +
  labs(title = "Population by Age Group", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

(p1 / p2) + plot_layout(heights = c(1.2, 1))
```

**Insights:**

1.  Most residents fall between **ages 25 to 54**
2.  **Youth population is shrinking**, suggesting long-term labor sustainability issues
3.  **Senior population (65+) rising**, indicating growing need for eldercare and aging population

### EDA 2: Gender Analysis

```{r fig.height=8, fig.width=10}
p3 <- ggplot(df, aes(x = Sex, y = Pop, fill = Sex)) +
  stat_summary(fun = sum, geom = "bar") +
  labs(title = "Population by Gender", x = NULL, y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

p4 <- df %>%
  group_by(Sex, AgeGroup) %>%
  summarise(Pop = sum(Pop)) %>%
  ggplot(aes(x = AgeGroup, y = Pop, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Gender Distribution by Age Group", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

(p3 / p4) + plot_layout(heights = c(1.2, 1))
```

**Insights:**

1.  Gender balance is nearly equal overall
2.  **Female dominates** in the senior age group, likely due to higher life expectancy

### EDA 3:

```{r fig.height=10, fig.width=10}
df_pyramid <- df %>%
  filter(AgeNum <= 90) %>%
  mutate(Pop = ifelse(Sex == "Males", -Pop, Pop))

p6 <- ggplot(df_pyramid, aes(x = AgeNum, y = Pop, fill = Sex)) +
  geom_col(width = 1) +
  coord_flip() +
  labs(title = "Population Pyramid", x = "Age", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

p7 <- df %>%
  group_by(Sex, AgeGroup) %>%
  summarise(Pop = sum(Pop)) %>%
  ggplot(aes(x = AgeGroup, y = Pop, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Age Group Distribution by Gender", y = "Population") +
  scale_y_continuous(labels = label_comma()) +
  common_theme

(p6 / p7) + plot_layout(heights = c(1.3, 1))
```

**Insights:**

1.  Pyramid shows **narrowing base wider top**, typical for aging societies
2.  **Adults dominate** across both genders, seniors are the second largest group

## 4. Conculsion

1.  Singapore faces a demographic shift towards aging, requiring proactive planning
2.  Uneven population spread across subzones and planning ares calls for smart urban development
3.  This EDA provides clear insights and serves as baseline for policy design, urban planning, and future modelling
